Seagrass mapping and monitoring \\ along the coasts of Crete, Greece

### 2011-02-07 MScDiplom_Lemenkova.tex
Valentijn's comments of 6Feb: 2(chart with buoys),4(broadband sensors); written 'Conclusions', updated 'Supervisory Committee'

\begin{itemize}
	\item[*] \href{http://www.southampton.ac.uk/}{University of Southampton}, \href{http://www.southampton.ac.uk/geography/}{School of Geography} (UK)\\
	\item[*] \href{http://www.lunduniversity.lu.se/}{Lund University}, \href{http://www.giscentrum.lu.se/english/index.htm}{GIS Centre} (Sweden)\\
	\item[*] \href{http://www.uw.edu.pl/en/}{University of Warsaw} (Poland)\\
	\item[*] \href{http://www.utwente.nl/en}{University of Twente}, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)} \\ (The Netherlands)
\end{itemize}

@@

Thesis submitted to the University of Twente, Faculty ITC, in partial fulfillment of the requirements
for the degree of Master of Science in Geo-information Science and Earth Observation for
Environmental Modelling and Management.\\

\textbf{Thesis Assessment Board}:
\begin{itemize}
	\item Name Examiner 1 (Chair)\dotfill
	\item Name Examiner 2\dotfill
	\item External Examiner \dotfill \\ \hfill{Dr.{\textit{Petter Pilesj\"o}\index[people]{Pilesj\"o, P.}} (Lund University, GIS Centre)}
\end{itemize}

@@

\textbf{Member Supervisors}: \\
\begin{itemize}
	\item[* ] Mr. \textit{Valentijn Venus}\index[people]{Venus, V.} \dotfill \\ (Department of Natural Resource, ITC)
	\item[* ] Dr. \textit{Albertus G. Toxopeus}\index[people]{Toxopeus, A.G.} \dotfill \\ (Department of Natural Resource, ITC)
\end{itemize}

@@

\textit{Chapter 5}: \nameref{sec:5} summarizes and briefly discusses again the main results, reported in detail in a previous chapter. This chapter proposes some final discussions to the reader, and comes to the main research outcomes.\\
Finally, we suggest some recommendations for any further research focusing on the seagrass environment along the coasts of Crete (or other Mediterranean areas) in the final \nameref{sec:6}.\\

@@

\subsubsection{Fieldwork equipment}
The research materials and equipment were provided by \href{http://147.52.130.151/MFIK_en/}{the Natural History Museum of Crete}, the \href{http://www.uoc.gr/Department/index.html}{University of Crete} and the ITC, and included the following items (Fig.\ref{fig:3.3}): 1) three iPAQs (Fig.\ref{fig:3.3a}); 2)
three GPS (Fig.\ref{fig:3.3c}); 3) Three underwater video cameras, \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d}), suitable for
photographing up to up to 33 foot depths and high-resolution 12.0-Mpixel image sensor; 4)
Markers and cords for the depths measurements; 5) Waterproof plastic Otterbox  (Fig.\ref{fig:3.3b})for keeping the
iPAQ dry; 6) \ac{SCUBA} Diving equipment, taken from the Ligaria Diving Center; 7) Boat (Fig.\ref{fig:3.4});

\begin{figure}[H]
	\centering
	\subfloat [iPAQ, HP]{\label{fig:3.3a}\includegraphics[width=0.3\textwidth]{Fig-19a.jpg}}
	\subfloat [Waterproof Otterbox for iPAQ]{\label{fig:3.3b}\includegraphics[width=0.3\textwidth]{Fig-19b.jpg}}
	\subfloat [GPS]{\label{fig:3.3c}\includegraphics[width=0.15\textwidth]{Fig-19c.jpg}}
	\subfloat [Olympus waterproof camera]{\label{fig:3.3d}\includegraphics[width=0.25\textwidth]{Fig-19d.jpg}}
	\caption{Fieldwork equipment}
	\label{fig:3.3}
\end{figure}

@@

\subsubsection{Sampling design}
The sampling design of the fieldwork was aimed at surveying of the spatial distribution of the
meadows of \textit{P.oceanica}, and spatial pattern of the seagrass meadows consisted from separate
patches. The fieldwork included several routes of the boat in the Ligaria beach sampling site, nine
routes in total, in the directions parallel to the coastline, ca 180-200 m long each one, thus enabling the
course plot to cover the area of growing seagrass: shelf areas not deeper than four meters.
The measurements of the seafloor cover types have been made using underwater video cameras
Olympus ST 8010, mounted under the boat to capture video footage and imagery. The data records
were made along each path using these cameras. A videographic approach, tested previously \cite{Norris97}\label{Norris97},
 was applied during the fieldwork on Crete island, for collecting information on benthic cover types and
distribution of the seagrass patches from photo transects, in order to use for the calibration of mapping approaches.
Three underwater video cameras, located on the bottom of the boat, provided videometric measurements of the
seafloor during the track, and resulted in a series of consequent overlapping images of the sea bottom under the
boat path. The general locations of the sampling sites and routes were selected on the basis of the visual examination
of the seagrass beds during snorkeling and \ac{SCUBA} diving, recommendations from the Greek collaborators of the \textit{Natural History Museum of Crete}, and available maps covering the research area. To ensure even and objective selection of sampling sites
we used transect sampling method, i.e. photographs were taken along the research path.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.70]{Fig-18.jpg}
	\caption{Boat used for the fieldwork measurements on Crete Island. 2010}
	\label{fig:3.4}
\end{figure}

Transect method is a common sampling technique in studies of the seagrass monitoring \cite{Shortis07}\label{Shortis07},
enabling the occurrences of the seagrass meadows to be recorded and counted in a systematic and
accurate way, detect spatial distribution of the single seagrass mattes to be properly identified,
without location bias. 
The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d}), at
five minutes interval along the tracklog path (Fig.\ref{fig:3.6}).
research sampling design also included measurements of depths (Fig. \ref{fig:A.4}), because bathymetry is one of the
most determining factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, \ac{iPAQ}, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The \ac{GPS} allowed to capture measurement locations on the \ac{iPAQ},
encapsulated in a plastic waterproof Otterbox (Fig \ref{fig:3.3b}). The camera were adjusted horizontally by a leveler and
 mounted under the bottom of the boat to capture video footage and imagery (Fig.\ref{fig:3.5}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state.

\begin{figure}[H]
	\centering
	\subfloat [view from below]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20a.jpg}}
	\subfloat [view \textit{en-face}]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20b.jpg}}
	\subfloat [view from above]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20c.jpg}}
	\caption{Scheme of placements of the Olimpus cameras during the measurements}
	\label{fig:3.5}
\end{figure}
The locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 m2 size each (Fig \ref{fig:A.7}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.40]{Fig-21.jpg}
	\caption{Locations of the video measurements and GPS tracklogs, Ligaria}
	\label{fig:3.6}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\subfloat [depth of 0.5 m]{\label{fig:32a}\includegraphics[width=0.4\textwidth]{Fig-32a.jpg}}
	\subfloat [depth of 2.0 m]{\label{fig:32b}\includegraphics[width=0.4\textwidth]{Fig-32b.jpg}}\\
	\subfloat [depth of 3.5 m]{\label{fig:32c}\includegraphics[width=0.4\textwidth]{Fig-32c.jpg}}
	\subfloat [depths of 0.5, 2.0 and 3.5 m]{\label{fig:32d}\includegraphics[width=0.4\textwidth]{Fig-32d.jpg}}
	\caption{Remote sensing reflectance of \textit{P.oceanica} at various depths. MERIS sensor.}
	\label{fig:4.9}
\end{figure}

@@

\section{Discussion and conclusions}\label{sec:5}
More than 50 \% of the world population lives within one km of the coast, which results in continued anthropogenic pressure on the coastal regions. Therefore, management of coastal resources and shelf zone protection become increasingly important nowadays, and require large-scale monitoring and mapping of the shelf areas as vital instrument for the environmental assessment. Besides Mediterranean area, the suggested methodology of the analysis of seagrass monitoring and analysis of itsadaptation towards different environmental conditions within the coastal zones can be applied to other marine coastal areas with growing seagrass.

\section*{Recommendations}\label{sec:6}
\markright{Recommendations}
\addcontentsline{toc}{section}{Recommendations}

@@

\subsection{\textit{Appendix-4}. Sampling design of the depths measurements. Ligaria beach, Crete}
%\appendixname{\textbf{4: \\Sampling design of the depths measurements. Ligaria beach, Crete}}

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.35]{Sampling_design.jpg}
		\caption{Scheme of the depths measurements; fieldwork on Ligaria beach}\label{fig:A.4}
	\end{center}	
\end{figure}
\pagebreak

\subsection{\textit{Appendix-5}. Results of videographic measurements: seafloor types on Crete island}
%\appendixname{\textbf{5: \\Results of videographic measurements: seafloor types on Crete island}}

@@

\subsection{\textit{Appendix-6}. Data pre-processing}
%\appendixname{\textbf{6: \\Data pre-processing}}

@@

\subsection{\textit{Appendix-7}. Data processing and statistical evaluation: Results of the measurements of the seafloor optical properties.}
%\appendixname{\textbf{7:\\ Data processing and statistical evaluation: \\ Results of the measurements of the seafloor optical properties.}}

@@

\subsection{\textit{Appendix-8}. Satellite and aerial images covering research area of Crete island: selected examples}
%\appendixname{\textbf{8:\\ Satellite and aerial images covering research area of Crete island: selected examples}}

@@

\subsection{\textit{Appendix-9}. Analysis and classification of the satellite and aerial images}

@@

\subsection{\textit{Appendix-10}. Research general workflow}
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.57]{UML_GIS.png}
	\caption{General methodological research approach. \href{http://inkscape.org/}{Inkscape}}
	\label{fig:1.11}
\end{figure}

>>>>

\begin{itemize}
	\item[*] \href{http://www.southampton.ac.uk/}{University of Southampton}, \href{http://www.southampton.ac.uk/geography/}{School of Geography} (UK)\\
	\item[*] \href{http://www.lunduniversity.lu.se/}{Lund University}, \href{http://www.giscentrum.lu.se/english/index.htm}{GIS Centre} (Sweden)\\
	\item[*] \href{http://www.uw.edu.pl/en/}{University of Warsaw}, \href{http://www.wgsr.uw.edu.pl/index.php?option=com_content&task=category&sectionid=10&id=32&Itemid=103}{Faculty of Geography and Regional Studies} (Poland)\\
	\item[*] \href{http://www.utwente.nl/en}{University of Twente}, \href{http://www.itc.nl/}{Faculty of Geo-Information Science and Earth Observation (ITC)} \\ (The Netherlands)
\end{itemize}

@@

Thesis submitted to the \textit{University of Twente, Faculty of Geo-Information Science and Earth Observation (ITC)}, in partial fulfillment of the requirements
for the degree of Master of Science in Geo-information Science and Earth Observation for
Environmental Modelling and Management.\\

\textbf{Thesis Assessment Board}:
\begin{itemize}
	\item Name Examiner 1 (Chair) \hfill{Prof. Dr. \textit{Andrew K. Skidmore}\index[people]{Skidmore A.K.} (ITC, Department of Natural Resource)}
	\item Name Examiner 2 \hfill{Ms. Ir. \textit{Louise M. van Leeuwen}\index[people]{Leeuwen, L. M. van} (ITC, Department of Natural Resource)}
	\item External Examiner \hfill{Dr. {\textit{Ma\l{}gorzata Roge-Wi\'sniewska}\index[people]{Roge-Wi\'sniewska, M.}} (University of Warsaw, Faculty of Geography and Regional Studies)}
\end{itemize}

@@

\textbf{Member Supervisors}: \\
\begin{itemize}
	\item[* ] Mr. \textit{Valentijn Venus}\index[people]{Venus, V.}  \hfill{(University of Twente, ITC, Department of Natural Resource)}
	\item[* ] Dr. \textit{Albertus G. Toxopeus}\index[people]{Toxopeus, A.G.}  \hfill{(University of Twente, ITC, Department of Natural Resource)}
\end{itemize}

@@

\textit{Cahpter 5}: \nameref{sec:5} discusses research results, reported in detail in a previous chapter, comes to the research insights and discusses relevance of this work in scientific context.\\
\textit{Chapter 6}: \nameref{sec:6} summarizes and highlights the final research outcomes and main results.\\
Finally, we suggest some recommendations for any further research focusing on the seagrass environment along the coasts of Crete (or other Mediterranean areas) in the final \nameref{sec:7}.\\

@@

\subsubsection{Fieldwork equipment}
The research materials and equipment were provided by \href{http://147.52.130.151/MFIK_en/}{the Natural History Museum of Crete}, the \href{http://www.uoc.gr/Department/index.html}{University of Crete} and the ITC, and included the following items (Fig.\ref{fig:3.3}): 1) three iPAQs (Fig.\ref{fig:3.3a}); 2)
three GPS (Fig.\ref{fig:3.3c}); 3) Three underwater video cameras, \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d}), suitable for photographing

\begin{figure}[H]
	\centering
	\subfloat [iPAQ, HP]{\label{fig:3.3a}\includegraphics[width=0.2\textwidth]{Fig-19a.jpg}}
	\subfloat [Waterproof Otterbox for iPAQ]{\label{fig:3.3b}\includegraphics[width=0.2\textwidth]{Fig-19b.jpg}}
	\subfloat [GPS]{\label{fig:3.3c}\includegraphics[width=0.15\textwidth]{Fig-19c.jpg}}
	\subfloat [Olympus waterproof camera]{\label{fig:3.3d}\includegraphics[width=0.2\textwidth]{Fig-19d.jpg}}
	\caption{Fieldwork equipment}
	\label{fig:3.3}
\end{figure}

up to up to 33 foot depths and high-resolution 12.0-Mpixel image sensor; 4)
Markers and cords for the depths measurements; 5) Waterproof plastic Otterbox  (Fig.\ref{fig:3.3b})for keeping the
iPAQ dry; 6) \ac{SCUBA} Diving equipment, taken from the Ligaria Diving Center; 7) Boat (Fig.\ref{fig:3.4});

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.60]{Fig-18.jpg}
	\caption{Boat used for the fieldwork measurements on Crete Island. 2010}
	\label{fig:3.4}
\end{figure}

@@

\subsubsection{Sampling design}
The sampling design of the fieldwork was aimed at surveying of the spatial distribution of the
meadows of \textit{P.oceanica}, and spatial pattern of the seagrass meadows consisted from separate
patches. The fieldwork included several routes of the boat in the Ligaria beach sampling site, nine
routes in total, in the directions parallel to the coastline, ca 180-200 m long each one, thus enabling the
course plot to cover the area of growing seagrass: shelf areas not deeper than four meters.

\begin{figure}[H]
	\begin{center}
		\includegraphics[scale=0.35]{Sampling_design.jpg}
		\caption{Scheme of the depths measurements; fieldwork on Ligaria beach}\label{fig:A.4}
	\end{center}	
\end{figure}

The measurements of the seafloor cover types have been made using underwater video cameras
Olympus ST 8010, mounted under the boat to capture video footage and imagery. The data records
were made along each path using these cameras. A videographic approach, tested previously \cite{Norris97}\label{Norris97},
 was applied during the fieldwork on Crete island, for collecting information on benthic cover types and
distribution of the seagrass patches from photo transects, in order to use for the calibration of mapping approaches.
Three underwater video cameras, located on the bottom of the boat, provided videometric measurements of the
seafloor during the track, and resulted in a series of consequent overlapping images of the sea bottom under the
boat path. The general locations of the sampling sites and routes were selected on the basis of the visual examination
of the seagrass beds during snorkeling and \ac{SCUBA} diving, recommendations from the Greek collaborators of the \textit{Natural History Museum of Crete}, and available maps covering the research area. To ensure even and objective selection of sampling sites
we used transect sampling method, i.e. photographs were taken along the research path. 
Transect method is a common sampling technique in studies of the seagrass monitoring \cite{Shortis07}\label{Shortis07},
enabling the occurrences of the seagrass meadows to be recorded and counted in a systematic and
accurate way, detect spatial distribution of the single seagrass mattes to be properly identified,
without location bias.

\begin{figure}[H]
	\centering
	\subfloat [view from below]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20a.jpg}}
	\subfloat [view \textit{en-face}]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20b.jpg}}
	\subfloat [view from above]{\label{fig:22a}\includegraphics[width=0.3\textwidth]{Fig-20c.jpg}}
	\caption{Scheme of placements of the Olimpus cameras during the measurements}
	\label{fig:3.5}
\end{figure}

The videographic measurements and photos, enabling to detect types of the
seafloor, were captured by means of three underwater digital video cameras \ac{Olympus ST} 8000  (Fig.\ref{fig:3.3d}), at
five minutes interval along the tracklog path (Fig.\ref{fig:3.6}).
research sampling design also included measurements of depths (Fig. \ref{fig:A.4}), because bathymetry is one of the
most determining factors for the seagrass locations. Measurements of depths were performed during
the fieldwork using cord, \ac{iPAQ}, and markers in order to assure that the videometric measurements
are taken at depths not more than four meters, in the shelf area.
As a result, nine transects were established of one m wide and 20 m long track, to cover seagrass beds
with videographic measurements. The \ac{GPS} allowed to capture measurement locations on the \ac{iPAQ},
encapsulated in a plastic waterproof Otterbox (Fig \ref{fig:3.3b}). The camera were adjusted horizontally by a leveler and
 mounted under the bottom of the boat to capture video footage and imagery (Fig.\ref{fig:3.5}). The data
were taken at proper weather conditions: sunny, serene and cloud-free days with glassy sea state. 
The locations of the route were randomly selected in the areas of the Ligaria beach, to ensure most dense
coverage of the seagrass meadows in the research area. The underwater measurements of the seagrass
coverage were carried out by taking video footage and photos of ca 0.5 m2 size each (Fig \ref{fig:A.7}).
The results of the underwater videometric measurements include series of digital images helping to
classify seafloor cover types and seagrass meadows, according to the differences in the structure,
colour, texture and shapes of the depicted objects.

\begin{figure}[H]
	\centering
	\includegraphics[scale=0.50]{Fig-21.jpg}
	\caption{Locations of the video measurements and GPS tracklogs, Ligaria}
	\label{fig:3.6}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\subfloat [depth of 0.5 m]{\label{fig:32a}\includegraphics[width=0.4\textwidth]{Fig-32a.jpg}}
	\subfloat [depth of 2.0 m]{\label{fig:32b}\includegraphics[width=0.4\textwidth]{Fig-32b.jpg}}\\
	\subfloat [depth of 3.5 m]{\label{fig:32c}\includegraphics[width=0.4\textwidth]{Fig-32c.jpg}}
	\subfloat [depths of 0.5, 2.0 and 3.5 m]{\label{fig:32d}\includegraphics[width=0.4\textwidth]{Fig-32d.jpg}}
	\caption{Remote sensing reflectance of \textit{P.oceanica} at various depths, simulated for broadband sensors.}
	\label{fig:4.9}
\end{figure}

@@

\section{Discussion}\label{sec:5}

\section{Conclusion}\label{sec:6}
The goal of this MSc research was to explore the perspectives, advantages and limitations of the narrow-band and broadband sensors for the environmental mapping and monitoring of \textit{P.oceanica} seagrass along the coasts of Crete Island. The research outcome demonstrated that the application of the remote sensing data from the broadband sensors is highly advantageous for the seagrass mapping, the spectral discrimination of \textit{P.oceanica} from other seafloor cover types is possible at diverse and changing environmental conditions, and that \textit{P.oceanica} is spectrally distinct from other seagrass species (\textit{Thalassia testudinum}).\vspace{1ex}\\
The \ac{RTM} software is a powerful means for analyzing spectral signatures of various seafloor types and enabling simulations of data received from broadband and narrowband remote sensors.\\ The example of application of \ac{WASI} \ac{RTM}, given in this work, is an achievement of the research insight towards the spectral properties of \textit{P.oceanica} and other bottom cover types, enabling to discriminate them from each other with changing environmental conditions. The research shows that spectral signatures of \textit{P.oceanica} are distinct at various depths.\vspace{1ex}\\
The methodology of the spectral discrimination of seafloor cover types is designed in the frame of this research and is based on the application of the remote sensing \ac{RTM} techniques, data from broadband sensors, hyperspectral radiometers for measurements of optical properties of the seawater, categorical and continuous statistical analysis for the data processing and \ac{GIS} raster based software for images visualization, classification and analysis. Technically, we used different software, adjusted for diverse research purposes, to manage, integrate and process data from various origin and resources, and finally to receive  accurate research results.\vspace{1ex}\\
The marine coastal ecosystems are complex, constantly changing and developing. Using flexibility of \ac{GIS} combined with remote sensing methods and application of data from broadband sensors is therefore advantageous for the monitoring of coastal areas. Besides Mediterranean area, the methodology of the seagrass environmental studies can be applied towards other shelf areas with dominating seagrass landscapes.\vspace{1ex}\\
More than 50 \% of the world population lives within one km of the coast, which results in continued anthropogenic pressure on the coastal regions. Therefore, management of coastal resources and shelf zone protection become increasingly important nowadays, and require large-scale monitoring and mapping of the shelf areas as a vital instrument for the environmental assessment.\vspace{1ex}\\
This research is a contribution to the development of the methodology of seagrass mapping with the aim of the environmental monitoring, and a case study of \textit{P.oceanica}, an endemic Mediterranean seagrass species, dominating in underwater landscapes and ecosystems along the coasts of Crete Island.
\pagebreak

\section*{Recommendations}\label{sec:7}
\markright{Recommendations}
\addcontentsline{toc}{section}{Recommendations}

@@

\subsection{\textit{Appendix-4}. Results of videographic measurements: seafloor types on Crete island}
%\appendixname{\textbf{5: \\Results of videographic measurements: seafloor types on Crete island}}

@@

\subsection{\textit{Appendix-5}. Data pre-processing}
%\appendixname{\textbf{6: \\Data pre-processing}}

@@

\subsection{\textit{Appendix-6}. Data processing and statistical evaluation: Results of the measurements of the seafloor optical properties.}
%\appendixname{\textbf{7:\\ Data processing and statistical evaluation: \\ Results of the measurements of the seafloor optical properties.}}

@@

\subsection{\textit{Appendix-7}. Satellite and aerial images covering research area of Crete island: selected examples}
%\appendixname{\textbf{8:\\ Satellite and aerial images covering research area of Crete island: selected examples}}

@@

\subsection{\textit{Appendix-8}. Analysis and classification of the satellite and aerial images}

@@

\subsection{\textit{Appendix-9}. Research general workflow}
\begin{figure}[H]
	\centering
	\includegraphics[scale=0.57]{UML_GIS.png}
	\caption{General methodological research approach. \href{http://inkscape.org/}{Inkscape}}
	\label{fig:1.11}
\end{figure}
